Lichat, A Lightweight Chat Protocol

### 2022-01-25 els2022-lichat/paper.tex
More fixes

\section{Related Work}\label{relatedwork}
Penpoppichanan et al.\cite{penpoppichanan2001security} identify a number of security and privacy related issues with the IRC protocol. While Lichat shares the basic concerns about privacy, being an unencrypted protocol, it does fix several of the identity and authentication related issues identified in the thesis.

The XMPP core specification\cite{saint2004extensible} alone is over 200 pages long, and does not include the already non-trivial question of parsing XML, which is used for its wire format. Saint-Andre et al.\cite{saint2009xmpp} also point out the lack of a proper end-to-end encryption scheme for the protocol.

@@

\section{Initial Design}\label{initial-design}
As a basis for the protocol we chose to use UTF-8 character streams for the transport. This ensures that protocol messages can be read and written by a human, which eases development and debugging. Specifying the character encoding also avoids ambiguity when interpreting the underlying bytes. For the actual data format, we relied on Common Lisp s-expressions, which could simply be \code{read} and \code{print}ed using the native Lisp functions.

@@

The client should also include a \code{:clock} field of the local Universal-Time at the time of sending. This ensures that, if distributed to other users, they can get an accurate ordering of events that reflects the sending user's point of view, without potential network delays changing the sequence of events. It also allows the server and clients to estimate lag in the network and issue warnings on unstable connections, or terminate them early, ensuring an ultimately more stable communication.

@@

On the client to client communication side, we decided to enforce all communication between clients to occur over channels. This simplifies the design on both sides, as there don't need to be any special rules to deal with direct communication. With all of this groundwork laid, we could implement a basic chat server that allowed for communication over multiple channels, as shown in \autoref{lst:basic-exchange}.

@@

After this most of the protocol design revolved around the implementation of a permissions system to allow channel operators better control over what users are allowed to do, and the ratification of distinct error types for potential failures.

The permissions system in general is very simple: each channel holds a map where each update type is associated with either a list of permitted users, or a list of denied users. As all communication between users must occur through channels, this provides a simple, but expressive model to handle moderation. For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.

@@

Further, we required the use of a null byte as an end of message marker. This is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length. We also specified that an optional, missing field is equivalent to specifying the field to be \code{nil}, and vice-versa. This simplifies initialising and dealing with optional values a lot.

To ease the initiation and maintenance of the connection, the \code{:from} and \code{:clock} fields were made optional, instead requiring the server to automatically fill them in should they be missing. This makes it even more trivial to write a simple, first client.

@@

Finally, we ratified the definition of all message types in a machine-readable format (\autoref{lst:spec}) that can be parsed using the same parser of the wire format. This allows protocol implementations to easily parse up-to-date representations of the entire message type hierarchy.

@@

JavaScript does not support multiple inheritance out of the box, so we implemented our own scheme following the serialised inheritance model of CLOS. This does however still not integrate neatly with the native JavaScript type system, meaning we also require the use of a specialised type test for dispatch.

@@

Our Elixir server implementation eschews inheritance altogether, as the language does not share an object-oriented system at all. We instead manually flatten the inheritance by specifying all inherited fields on each message type. Each message type is represented through a module, with a function to handle the update on the server-side. Thanks to Elixir's macro system we were able to avoid most of the boilerplate overhead of doing this, but we have so far not implemented automated creation of the message representations from the machine-readable specification.

@@

A particular lament with IRC is the lack of easy file sharing. Users have come to expect being able to share images, videos, and other small files directly through the chat platform. IRC does have the DCC (Direct Client to Client) protocol extension, but it only allows exchanging files between two users directly, rather than broadcasting it to a channel. The ``shirakumo-data'' extension introduces a new message type that allows sending a base64-encoded payload, with an associated mime-type. The server can restrict permitted mime-types, and the general message length checking (with the null-byte skipping mechanism) allows dropping data payloads that are too large. The ``shirakumo-link'' extension further optimises this by letting the server save data-payloads and distribute them via an internet link instead, reducing the amount of data needing to be sent out to the channel drastically.

@@

>>>>

\section{Related Work}\label{relatedwork}
Penpoppichanan\cite{penpoppichanan2001security} identifies a number of security and privacy related issues with the IRC protocol. While Lichat shares the basic concerns about privacy, being an unencrypted protocol, it does fix several of the identity and authentication related issues identified in the thesis.

The XMPP core specification\cite{saint2004extensible} alone is over 200 pages long, and does not include the already non-trivial question of parsing XML, which is used for its wire format. Saint-Andre\cite{saint2009xmpp} also points out the lack of a proper end-to-end encryption scheme for the protocol.

@@

\section{Initial Design}\label{initial-design}
As a basis for the protocol we chose to use UTF-8 character streams for the transport. This choice ensures that protocol messages can be read and written by a human, which eases development and debugging. Specifying the character encoding also avoids ambiguity when interpreting the underlying bytes. For the actual data format, we relied on Common Lisp s-expressions, which could simply be \code{read} and \code{print}ed using the native Lisp functions.

@@

The client should also include a \code{:clock} field of the local Universal-Time at the time of sending. This field ensures that, if distributed to other users, they can get an accurate ordering of events that reflects the sending user's point of view, without potential network delays changing the sequence of events. It also allows the server and clients to estimate lag in the network and issue warnings on unstable connections, or terminate them early, ensuring an ultimately more stable communication.

@@

On the client to client communication side, we decided to enforce all communication between clients to occur over channels. This decision simplifies the design on both sides, as there don't need to be any special rules to deal with direct communication. With all of this groundwork laid, we could implement a basic chat server that allowed for communication over multiple channels, as shown in \autoref{lst:basic-exchange}.

@@

After these changes most of the protocol design revolved around the implementation of a permissions system to allow channel operators better control over what users are allowed to do, and the ratification of distinct error types for potential failures.

The permissions system in general is very simple: each channel holds a map where each update type is associated with either a list of permitted users, or a list of denied users. As all communication between users must occur through channels, this system provides a simple, but expressive model to handle moderation. For instance, banning a user simply requires removing their ability to send a \code{join} message to the channel.

@@

Further, we required the use of a null byte as an end of message marker. The null byte is useful so that a server or client can restore the stream of messages should a parser failure occur, or should the update be rejected due to excessive length. We also specified that an optional, missing field is equivalent to specifying the field to be \code{nil}, and vice-versa. This aliasing simplifies initialising and dealing with optional values a lot.

To ease the initiation and maintenance of the connection, the \code{:from} and \code{:clock} fields were made optional, instead requiring the server to automatically fill them in should they be missing. This behaviour makes it even more trivial to write a simple, first client.

@@

Finally, we ratified the definition of all message types in a machine-readable format (\autoref{lst:spec}) that can be parsed using the same parser of the wire format. These machine-readable specs allow protocol implementations to easily parse up-to-date representations of the entire message type hierarchy.

@@

JavaScript does not support multiple inheritance out of the box, so we implemented our own scheme following the serialised inheritance model of CLOS. This scheme does however still not integrate neatly with the native JavaScript type system, meaning we also require the use of a specialised type test for dispatch.

@@

Our Elixir server implementation eschews inheritance altogether, as the language does not share an object-oriented system at all. We instead manually flatten the inheritance by specifying all inherited fields on each message type. Each message type is represented through a module, with a function to handle the update on the server-side. Thanks to Elixir's macro system we were able to avoid most of the boilerplate overhead of writing each message type, but we have so far not implemented automated creation of the message representations from the machine-readable specification.

@@

A particular lament with IRC is the lack of easy file sharing. Users have come to expect being able to share images, videos, and other small files directly through the chat platform. IRC does have the DCC (Direct Client to Client) protocol extension, but it only allows exchanging files between two users directly, rather than broadcasting it to a channel. The ``shirakumo-data'' extension introduces a new message type that allows sending a base64-encoded payload, with an associated mime-type. The server can restrict permitted mime-types, and the general message length checking (with the null-byte skipping mechanism) allows dropping data payloads that are too large. The ``shirakumo-link'' extension further optimises sending media by letting the server save data-payloads and distribute them via an internet link instead, reducing the amount of data needing to be sent out to the channel drastically.

@@

\section{Source Code}\label{source}
The lichat specification and implementations can be found freely on the web:
\begin{itemize}
\item Lichat \url{https://shirakumo.org/projects/lichat}
\item JavaScript \url{https://shirakumo.org/projects/lichat-js}
\item Elixir \url{https://shirakumo.org/projects/ex-lichat}
\item Java \url{https://shirakumo.org/projects/jlichat}
\item Python \url{https://shirakumo.org/projects/py-lichat}
\end{itemize}
